Staged Abstract Interpreters

### 2019-09-07 paper/10_intro.tex
typo

Based on the idea of AAM, more pragmatically, several implementation strategies
in high-level functional programming languages have emerged.  Such techniques
include monads and monad transformers \cite{DBLP:journals/pacmpl/DaraisLNH17,
Sergey:2013:MAI:2491956.2491979}, arrows
\cite{Keidel:2018:CSP:3243631.3236767}, extensible effects
\cite{Kiselyov:2015:FMM:2804302.2804319, Githubsemantic}, etc. These pure
approaches provide certain benefits, for example, the abstract interpretation
artifacts can be built in a compositional and modular fashion, and referential
transparency allows programmers to confidently reason about the correctness of
their implementations.  The soundness of an analysis can also be proven with less
effort, whether by mechanized proofs \cite{Darais:2016:CGC:2951913.2951934} or
paper-based proofs \cite{Keidel:2018:CSP:3243631.3236767}.

>>>>

Based on the idea of AAM, more pragmatically, several implementation strategies
in high-level functional programming languages have emerged.  Such techniques
include monads and monad transformers \cite{DBLP:journals/pacmpl/DaraisLNH17,
Sergey:2013:MAI:2491956.2491979}, arrows
\cite{Keidel:2018:CSP:3243631.3236767}, extensible effects
\cite{Kiselyov:2015:FMM:2804302.2804319, Githubsemantic}, etc. These pure
approaches provide certain benefits, for example, the abstract interpretation
artifacts can be built in a compositional and modular fashion, and referential
transparency allows programmers to confidently reason about the correctness of
their implementations.  The soundness of an analysis can also be proven with less
effort, either by mechanized proofs \cite{Darais:2016:CGC:2951913.2951934} or
paper-based proofs \cite{Keidel:2018:CSP:3243631.3236767}.

### 2019-09-07 paper/40_absinterp.tex
typo

After seeing the unstaged and staged concrete interpreter, now we turn our
focus to abstract interpreters under the same framework. We first present a
lattice representation with binding-time abstraction, as well as simple
abstract domains, such as power set lattice.  Then we show the abstract
components, namely abstract values, abstract environments, and abstract stores.
The abstract interpreter we construct in this section is similar to
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s. For simplicity, it is
context/path/flow-insensitive by our choice. In Section \ref{cfa}, we will
instantiate it to a commonly-used context-sensitive analysis;
\citet{Darais:2015:GTM:2814270.2814308} show how to achieve path- and
flow-sensitivity by varying the monads, which is also applicable to our
unstaged or staged abstract interpreter.

@@

The co-inductive caching algorithm is implemented as an instrumentation over the
@eval@ function (Figure \ref{fig:coind_cache}), and it also closes the open recursion.
The instrumentation works as follows. Initially, the two caches are both empty.
During the iteration, we first check whether @out@ contains the configuration
@cfg@, representing the current desired computation. If it does, then the result
is directly returned through the monad stack.
Otherwise, we first retrieve the result from @in@ (@⊥@ if @in@ not contains
@cfg@), and update the result into the @out@ cache in the fashion of join.
Then, we invoke @ev@ to compute the result for this iteration, where @ev@ takes
@fix(ev)@ as a self reference.
After the evaluation, the store @σ@ may have been changed. Therefore, 
we get the latest store and construct the result @(v, $\sigma$)@; the
result will be used to update the out cache.

>>>>

After seeing the unstaged and staged concrete interpreter, now we turn our
focus to abstract interpreters under the same framework. We first present a
lattice representation with binding-time abstraction, as well as simple
abstract domains, such as power sets.  Then, we show the abstract
components, namely, abstract values, abstract environments, and abstract stores.
The abstract interpreter we construct in this section is similar to
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s. For simplicity, it is
context/path/flow-insensitive by our choice. In Section \ref{cfa}, we will
instantiate it to a commonly-used context-sensitive analysis;
\citet{Darais:2015:GTM:2814270.2814308} show how to achieve path- and
flow-sensitivity by varying the monads, which is also applicable to our
unstaged or staged abstract interpreter.

@@

The co-inductive caching algorithm is implemented as an instrumentation over the
@eval@ function (Figure \ref{fig:coind_cache}), and it also closes the open recursion.
The instrumentation works as follows. Initially, the two caches are both empty.
During the iteration, we first check whether @out@ contains the configuration
@cfg@, representing the current desired computation. If it does, then the result
is directly returned through the monad stack.
Otherwise, we first retrieve the result from @in@ (@⊥@ if @in@ not contains
@cfg@), and update the result into the @out@ cache in the fashion of join.
Then, we invoke @ev@ to compute the result for this iteration, where @ev@ takes
@fix(ev)@ as a self reference.
After the evaluation, the store @σ@ may have been changed. Therefore, 
we get the latest store and construct the result @(v, σ)@; the
result will be used to update the out cache.

### 2019-09-07 paper/50_stageabs.tex
typo

In Section \ref{stagedpoly_lat}, we exploited the higher-kinded type @R@ to
achieve the stage polymorphism. Now we instantiate the type @R@ as @Rep@ and
still use power-set as example to describe its staged version.
\begin{lstlisting}[escapechar=!]
  trait RepLattice[S] extends Lattice[S, Rep]
  def RepSetLattice[T]: RepLattice[Set[T]] = new RepLattice[Set[T]] {
    lazy val ⊥: !\hl{Rep[Set[T]]}! = Set[T]()
    lazy val ⊤: !\hl{Rep[Set[T]]}! = error("No representation for ⊤")
    def ⊑(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Boolean]}! = l1 subsetOf  l2
    def ⊔(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 union     l2
    def ⊓(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 intersect l2
  }
\end{lstlisting}

@@

\paragraph{Staged Monads for Abstract Interpretation}
We use the same monad stack structure from the unstaged abstract interpreter,
but replacing \textit{all} the monad transformers with their staged versions. The
following code snippet shows this change. We manually fuse the three inner
transformers (@IdM@ omitted) into a single monad @RepSetReaderStateM[A, B, C]@,
where @A@ is the type parameter for the reader effects and @B@ for the state effects.
In our monad stack scheme, they are both instantiated with the type @Cache@. 
Similar to the unstaged version, the grounded result type is a pair of two
staged value: @Rep[Set[(Value, Store)]]@ and @Rep[Cache]@.
\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type AnsM[T] = RepReaderT[RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[Set[(Value, Store)]], Rep[Cache])
    ...
  }
\end{lstlisting}

@@

\paragraph{Partially-static Data}
Our treatment of binding-times is coarse-grained: @Expr@s is static, the rest
of the components are all dynamic. But this is not always true since the static
data has to be used somewhere with dynamic operations. Partial-static data is a
way to improve binding-times and optimize the generated code.

>>>>

In Section \ref{stagedpoly_lat}, we exploited the higher-kinded type @R@ to
achieve the stage polymorphism. Now we instantiate the type @R@ as @Rep@ and
still use the power-set as example to describe its staged version.
\begin{lstlisting}[escapechar=!]
  trait RepLattice[S] extends Lattice[S, Rep]
  def RepSetLattice[T]: RepLattice[Set[T]] = new RepLattice[Set[T]] {
    lazy val ⊥: !\hl{Rep[Set[T]]}! = Set[T]()
    lazy val ⊤: !\hl{Rep[Set[T]]}! = error("No representation for ⊤")
    def ⊑(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Boolean]}! = l1 subsetOf  l2
    def ⊔(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 union     l2
    def ⊓(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 intersect l2
  }
\end{lstlisting}

@@

\paragraph{Staged Monads for Abstract Interpretation}
We use the same monad stack structure from the unstaged abstract interpreter,
but replacing \textit{all} the monad transformers with their staged versions. The
following code snippet shows this change. We manually fuse the three inner
transformers (@IdM@ omitted) into a single monad @RepSetReaderStateM[A, B, C]@,
where @A@ is the type parameter for the reader effects and @B@ for the state effects.
In our monad stack scheme, they are both instantiated with the type @Cache@. 
Similar to the unstaged version, the grounded result type is a pair of two
staged values: @Rep[Set[(Value, Store)]]@ and @Rep[Cache]@.
\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type AnsM[T] = RepReaderT[RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[Set[(Value, Store)]], Rep[Cache])
    ...
  }
\end{lstlisting}

@@

\paragraph{Partially-static Data}
Our treatment of binding-times is coarse-grained: @Expr@s is static, the rest
of the components are all dynamic. But this is not always true, because the static
data have to be used somewhere with the dynamic operations. Partial-static data is a
way to improve binding-times and optimize the generated code.

### 2019-09-07 paper/80_related.tex
typo

\paragraph{Optimizing Static Analysis Through Specialization} The underlying
idea in this paper is closely related to abstract compilation (AC)
\cite{Boucher:1996:ACN:647473.727587}: removing the interpretative overhead of
traversing the syntax tree by specialization. The residual programs of AC can
be either textual (using quasi-quotations) or closures (using higher-order
functions).  Our approach is more flexible between representations and requires
less engineering efforts to refactor the analyzer (Section \ref{cs_ac}).
\citet{Johnson:2013:OAA:2500365.2500604} adapted closure generation to optimize
small-step abstract interpreter in state-transition style. The program
is first compiled to an ``abstract bytecode'' IR. The IR consists of higher-order
functions, which will be executed later on an abstract abstract machine for
that IR.  \citet{damian1999partial} provided a formal treatment to abstract
compilation and control-flow analysis, as well as proofs to establish the
correctness of a certified specialized analyzer.  \citet{amtoft1999partial}
applied partial evaluation for constraint-based control-flow analysis.

Splitting an analysis into multiple stages has also been studied for other
forms of static analyses.
\citet{DBLP:conf/cgo/HardekopfL11} applied staging to a
flow-sensitive pointer analysis. The first stage is to analyze the
programs to obtain a sparse representation, and then the second
stage conducts the flow-sensitive analysis by refining the first
one. Another large body of static analyses is implemented by
first generating constraints and then using a solver to discharge the
proof obligations or find the fixed-point. Notable examples include
SAT- and SMT-based program analysis
\cite{Gulwani:2008:PAC:1375581.1375616} and Datalog-based points-to
analysis \cite{Smaragdakis:2015:PA:2802194.2802195}.
Such analyses can be considered having stage distinctions.  One piece of future
work is to explore whether they can be implemented and optimized from
meta-programming.
Souffl{\'e}, one of the state-of-the-art Datalog engines commonly used in
points-to analysis \cite{Antoniadis:2017:PDS:3088515.3088522}, uses Futamura
projections for efficient implementation \cite{10.1007/978-3-319-41540-6_23}.
However, it only specializes Datalog rules to the running engine, instead of a
full pipeline from a program AST to a specialized solver.

Recent work shows that numerical abstract domains can be
specialized w.r.t. the static program structure, although not by
way of staging. For example, decomposing polyhedrons
\cite{DBLP:conf/popl/SinghPV17, Singh:2017:PCD:3177123.3158143} to
smaller ones depending on the variables involved in a statement can
significantly improve the running time and space consumption of abstract
transformers.

>>>>

\paragraph{Optimizing Static Analysis Through Specialization} The underlying
idea in this paper is closely related to abstract compilation (AC)
\cite{Boucher:1996:ACN:647473.727587}: removing the interpretative overhead of
traversing the syntax tree by specialization. The residual programs of AC can
be either textual (using quasi-quotations) or closures (using higher-order
functions).  Our approach is more flexible between representations and requires
less engineering efforts to refactor the analyzer (Section \ref{cs_ac}).
\citet{Johnson:2013:OAA:2500365.2500604} adapted closure generation to optimize a
small-step abstract interpreter in state-transition style. The program
is firstly compiled to an ``abstract bytecode'' IR. The IR consists of higher-order
functions, which will be executed later on an abstract abstract machine for
that IR.  \citet{damian1999partial} provided a formal treatment for abstract
compilation and control-flow analysis, as well as proofs to establish the
correctness of a certified specialized analyzer.  \citet{amtoft1999partial}
applied partial evaluation for constraint-based control-flow analysis.

Splitting an analysis into multiple stages has also been studied for other
forms of static analyses.
\citet{DBLP:conf/cgo/HardekopfL11} applied staging to a
flow-sensitive pointer analysis. The first stage is to analyze the
programs and obtain a sparse representation, and then the second
stage conducts a flow-sensitive analysis by refining the first
one. Another large body of static analyses is implemented by
first generating constraints and then using a solver to discharge the
proof obligations or find the fixed-point. Notable examples include
SAT- and SMT-based program analysis
\cite{Gulwani:2008:PAC:1375581.1375616} and Datalog-based points-to
analysis \cite{Smaragdakis:2015:PA:2802194.2802195}.
Such analyses can be considered having stage distinctions.  One piece of future
work is to explore whether they can be implemented and optimized using
meta-programming.
Souffl{\'e}, one of the state-of-the-art Datalog engines commonly used in
points-to analysis \cite{Antoniadis:2017:PDS:3088515.3088522}, uses Futamura
projections for efficient implementation \cite{10.1007/978-3-319-41540-6_23}.
However, it only specializes the solving engine to Datalog rules, instead of a
full pipeline from a program AST to a specialized solver.

Recent work shows that numerical abstract domains can be
specialized to the static program structure, although not by
ways of staging. For example, decomposing polyhedrons
\cite{DBLP:conf/popl/SinghPV17, Singh:2017:PCD:3177123.3158143} to
smaller ones depending on the variables involved in a statement can
significantly improve the running time and space consumption of abstract
transformers.
